% This is the Reed College LaTeX thesis template. Most of the work
% for the document class was done by Sam Noble (SN), as well as this
% template. Later comments etc. by Ben Salzberg (BTS). Additional
% restructuring and APA support by Jess Youngberg (JY).
% Your comments and suggestions are more than welcome; please email
% them to cus@reed.edu
%
% See http://web.reed.edu/cis/help/latex.html for help. There are a
% great bunch of help pages there, with notes on
% getting started, bibtex, etc. Go there and read it if you're not
% already familiar with LaTeX.
%
% Any line that starts with a percent symbol is a comment.
% They won't show up in the document, and are useful for notes
% to yourself and explaining commands.
% Commenting also removes a line from the document;
% very handy for troubleshooting problems. -BTS

% As far as I know, this follows the requirements laid out in
% the 2002-2003 Senior Handbook. Ask a librarian to check the
% document before binding. -SN

%%
%% Preamble
%%
% \documentclass{<something>} must begin each LaTeX document
\documentclass[12pt,twoside]{reedthesis}
% Packages are extensions to the basic LaTeX functions. Whatever you
% want to typeset, there is probably a package out there for it.
% Chemistry (chemtex), screenplays, you name it.
% Check out CTAN to see: http://www.ctan.org/
%%
\usepackage{graphicx,latexsym}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{longtable,booktabs,setspace}
\usepackage{chemarr} %% Useful for one reaction arrow, useless if
% you're not a chem major
\usepackage[hyphens]{url}
\usepackage{rotating}
% \usepackage{natbib}
% Comment out the natbib line above and uncomment the following two
% lines to use the new
% biblatex-chicago style, for Chicago A. Also make some changes at
% the end where the
% bibliography is included.
\usepackage{biblatex-chicago}
\bibliography{thesis}

% \usepackage{times} % other fonts are available like times, bookman,
% charter, palatino

\title{Empirical Analysis of Fair Hierarchical Clustering Algorithms}
\author{Param Kapur}
% The month and year that you submit your FINAL draft TO THE LIBRARY
% (May or December)
\date{December 2025}
\division{Mathematical and Natural Sciences}
\advisor{Harper Knittel}
%If you have two advisors for some reason, you can use the following
%\altadvisor{Your Other Advisor}
%%% Remember to use the correct department!
\department{Computer Science}
% if you're writing a thesis in an interdisciplinary major,
% uncomment the line below and change the text as appropriate.
% check the Senior Handbook if unsure.
%\thedivisionof{The Established Interdisciplinary Committee for}
% if you want the approval page to say "Approved for the Committee",
% uncomment the next line
%\approvedforthe{Committee}

\setlength{\parskip}{0pt}
%%
%% End Preamble
%%
%% The fun begins:
\begin{document}

\maketitle
\frontmatter % this stuff will be roman-numbered
\pagestyle{empty} % this removes page numbers from the frontmatter

% Acknowledgements (Acceptable American spelling) are optional
% So are Acknowledgments (proper English spelling)
\chapter*{Acknowledgements}
I want to thank a few people.

% The preface is optional
% To remove it, comment it out or delete it.
\chapter*{Preface}
This is an example of a thesis setup to use the reed thesis document class.

\chapter*{List of Abbreviations}
You can always change the way your abbreviations are formatted. Play
around with it yourself, use tables, or come to CUS if you'd like to
change the way it looks. You can also completely remove this chapter
if you have no need for a list of abbreviations. Here is an example
of what this could look like:

\begin{table}[h]
  \centering % You could remove this to move table to the left
  \begin{tabular}{ll}
    \textbf{ABC}    &  American Broadcasting Company \\
    \textbf{CBS}    &  Columbia Broadcasting System\\
    \textbf{CDC}    &  Center for Disease Control \\
    \textbf{CIA}    &  Central Intelligence Agency\\
    \textbf{CLBR}   &  Center for Life Beyond Reed\\
    \textbf{CUS}    &  Computer User Services\\
    \textbf{FBI}    &  Federal Bureau of Investigation\\
    \textbf{NBC}    &  National Broadcasting Corporation\\
  \end{tabular}
\end{table}

\tableofcontents
% if you want a list of tables, optional
\listoftables
% if you want a list of figures, also optional
\listoffigures

% The abstract is not required if you're writing a creative thesis
% (but aren't they all?)
% If your abstract is longer than a page, there may be a formatting issue.
\chapter*{Abstract}
The preface pretty much says it all.

\chapter*{Dedication}
You can have a dedication here if you wish.

\mainmatter % here the regular arabic numbering starts
\pagestyle{fancyplain} % turns page numbering back on

%The \introduction command is provided as a convenience.
%if you want special chapter formatting, you'll probably want to
% avoid using it altogether

\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}
\chaptermark{Introduction}
\markboth{Introduction}{Introduction}
% The three lines above are to make sure that the headers are right,
% that the intro gets included in the table of contents, and that it
% doesn't get numbered 1 so that chapter one is 1.

% Double spacing: if you want to double space, or one and a half
% space, uncomment one of the following lines. You can go back to
% single spacing with the \singlespacing command.
\onehalfspacing
% \doublespacing

Algorithmic development lies at the core of computer science, driving
innovations across virtually every technology and application domain.
Theoretical research provides essential insights, including
guarantees on correctness, computational complexity, and solution
quality. These theoretical properties guide the initial design and
selection of algorithms. Yet, translating these theoretical
guarantees into practical, real-world performance is often complex
and unpredictable. Real-world factors such as implementation
overhead, hardware-specific nuances, unique data characteristics, and
hidden constant factors can substantially alter an algorithm’s
observed performance compared to its theoretical expectations.

This disconnect—commonly referred to as the \emph{theory–practice
gap}—is particularly pronounced in areas that handle massive
datasets, require real-time or near-real-time responses, or involve
balancing multiple objectives such as fairness against traditional
performance metrics. Fair hierarchical clustering, for instance,
comes with theoretical bounds on the expected increase in clustering
cost due to fairness constraints; however, actual performance
trade-offs related to runtime and solution quality under realistic
conditions often remain unclear. Similarly, theoretical predictions
about parallel algorithms’ speedups can fail to materialize in
practice due to overlooked factors like communication overhead, while
approximation algorithms frequently outperform their theoretical
worst-case guarantees on typical problem instances.

Addressing this theory–practice gap is crucial because it influences
the adoption and practical impact of algorithms. While theory
provides essential worst-case or average-case assurances,
practitioners often lack reliable, evidence-based guidance to predict
real-world performance accurately. Consequently, this uncertainty can
either deter the use of theoretically superior algorithms or result
in suboptimal algorithmic choices based on anecdotal experiences or
incomplete testing. Despite its significance, systematic empirical
evaluation of algorithmic performance—especially in domains
exhibiting notable theory-practice discrepancies—is still rare.

This thesis tackles precisely this issue by providing a
comprehensive, structured empirical analysis designed to bridge the
existing theory–practice gap. Through an extensive literature review,
I first identify key algorithmic areas—such as fair hierarchical
clustering, parallel graph processing, and approximation methods—that
demonstrate significant divergence between theoretical predictions
and empirical observations. Recognizing the need for rigorous,
reproducible, and standardized testing, I have developed a robust
benchmarking framework specifically aimed at controlled
experimentation and comprehensive performance evaluations.

Utilizing this benchmarking environment, the research reported in
this thesis includes extensive empirical studies involving both
synthetic datasets—which probe specific theoretical
characteristics—and diverse real-world datasets that reflect typical
practical scenarios. These experiments systematically compare
theoretical metrics like runtime complexity, approximation ratios,
and fairness guarantees with empirical outcomes such as actual
execution times, memory consumption, solution quality, and fairness
metrics. Rather than isolated studies, this thesis presents an
integrated view, uncovering detailed patterns and insights into
algorithmic behavior under realistic conditions.

Ultimately, the contribution of this thesis lies in providing clear,
systematic empirical evidence to better inform algorithm selection
and implementation decisions, effectively bridging the gap between
theoretical expectations and practical realities.

\chapter{Background}
This is the first page of the first chapter. You may delete the
contents of this chapter so you can add your own text; it's just here
to show you some examples.

\section{Fairness Concepts in Machine Learning}\label{sec:fairness_concepts}
\input{sections/background/2.1-fairness-in-ml.tex}

\section{Preliminaries and Fairness in Clustering}\label{sec:fair_clustering}
\input{sections/background/2.2-fair-clustering.tex}

\section{Hierarchical Clustering and Cost}
\input{sections/background/2.3-hierarchical-clustering-and-cost.tex}

\section{Fair Hierarchical
Clustering}\label{subsec:fair_hierarchical_clustering}
\input{sections/background/2.4-fair-hierarchical-clustering.tex}

\input{sections/background/2.5-theory-practice-benchmarking.tex}

\chapter*{Conclusion}
\addcontentsline{toc}{chapter}{Conclusion}
\chaptermark{Conclusion}
\markboth{Conclusion}{Conclusion}
\setcounter{chapter}{4}
\setcounter{section}{0}

Here's a conclusion, demonstrating the use of all that manual
incrementing and table of contents adding that has to happen if you
use the starred form of the chapter command. The deal is, the chapter
command in \LaTeX\ does a lot of things: it increments the chapter
counter, it resets the section counter to zero, it puts the name of
the chapter into the table of contents and the running headers, and
probably some other stuff.

So, if you remove all that stuff because you don't like it to say
``Chapter 4: Conclusion'', then you have to manually add all the
things \LaTeX\ would normally do for you. Maybe someday we'll write a
new chapter macro that doesn't add ``Chapter X'' to the beginning of
every chapter title.

\section{More info}
And here's some other random info: the first paragraph after a
chapter title or section head \emph{shouldn't be} indented, because
indents are to tell the reader that you're starting a new paragraph.
Since that's obvious after a chapter or section title, proper
typesetting doesn't add an indent there.

%If you feel it necessary to include an appendix, it goes here.
\appendix
\chapter{The First Appendix}
\chapter{The Second Appendix, for Fun}

%This is where endnotes are supposed to go, if you have them.
%I have no idea how endnotes work with LaTeX.

\backmatter % backmatter makes the index and bibliography appear
% properly in the t.o.c...

% if you're using bibtex, the next line forces every entry in the
% bibtex file to be included
% in your bibliography, regardless of whether or not you've cited it
% in the thesis.
\nocite{*}

% Rename my bibliography to be called "Works Cited" and not
% "References" or ``Bibliography''
% \renewcommand{\bibname}{Works Cited}

%    \bibliographystyle{bsts/mla-good} % there are a variety of
% styles available;
%  \bibliographystyle{plainnat}
% replace ``plainnat'' with the style of choice. You can refer to
% files in the bsts or APA
% subfolder, e.g.
% \bibliographystyle{APA/apa-good}  % or
% \bibliography{thesis}
% Comment the above two lines and uncomment the next line to use
% biblatex-chicago.
\printbibliography[heading=bibintoc]

% Finally, an index would go here... but it is also optional.
\end{document}
